---
title: "Data Wrangling Project"
author: "David Crook"
date: "October 20, 2015"
output: 
  html_document:
    toc: true
    theme: united
---

***

### Setup

``` {r Initial setup}
setwd("~/projects/Classes/FoundationsOfDataScience_sliderule/github/part4_data_wrangling/project")

library(ggplot2)
suppressMessages(library(psych))
library(plyr)
suppressMessages(library(dplyr))
#library(stringr)

```

***

### Problem Description

You should create one R script called `run_analysis.R` that does the following.

1. ~~Merges the training and the test sets to create one data set.~~
1. ~~Extracts only the measurements on the mean and standard deviation for each measurement.~~
1. ~~Uses descriptive activity names to name the activities in the data set~~
1. ~~Appropriately labels the data set with descriptive variable names.~~
1. ~~From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.~~

 
*** 

### Data import

``` {r data_import, echo=FALSE}
# read in features and activity labels
features   <- read.csv ("UCI_HAR_Dataset/features.txt", sep = ' ', header = FALSE)
colnames(features) <- c("feature_column",  "feature_name")
str(features)
# 'data.frame':	561 obs. of  2 variables:
# $ V1: int  1 2 3 4 5 6 7 8 9 10 ...
#  $ V2: Factor w/ 477 levels "angle(tBodyAccJerkMean),gravityMean)",..: 243 244 245 250 251 252 237 238 239 240 ...

act_labels <- read.csv ("UCI_HAR_Dataset/activity_labels.txt", sep = ' ', header = FALSE)
colnames(act_labels) <- c("activity_class_encoding",  "activity_name")
act_labels$activity_class_encoding <- factor(act_labels$activity_class_encoding)
act_labels

# -----------------
# test set
# -----------------
# $ wc -l test/X_test.txt test/subject_test.txt test/y_test.txt
#     2947 test/X_test.txt
#     2947 test/subject_test.txt
#     2947 test/y_test.txt
    
test_act_labels  <- read.csv ("UCI_HAR_Dataset/test/y_test.txt", sep = ' ', header = FALSE)
dim(test_act_labels)

test_subj_labels  <- read.csv ("UCI_HAR_Dataset/test/subject_test.txt", sep = ' ', header = FALSE)
dim(test_subj_labels)

test_feature_measurements <- read.csv ("UCI_HAR_Dataset/test/X_test.txt", sep = '', header = FALSE)
dim(test_feature_measurements)

# describe(test_feature_measurements)

# check that activity levels have same number of observations as measurements
print(ifelse((dim(test_feature_measurements)[1] != dim(test_act_labels)[1]) &
             (dim(test_feature_measurements)[1] != dim(test_subj_labels)[1]),
             "Error: length mismatch in test data",
             "Good: length match in test data"))

# -----------------
# train set
# -----------------
# wc -l train/X_train.txt train/subject_train.txt train/y_train.txt
#     7352 train/X_train.txt
#     7352 train/subject_train.txt
#     7352 train/y_train.txt    
    
train_act_labels  <- read.csv ("UCI_HAR_Dataset/train/y_train.txt", sep = ' ', header = FALSE)
dim(train_act_labels)

train_subj_labels  <- read.csv ("UCI_HAR_Dataset/train/subject_train.txt", sep = ' ', header = FALSE)
dim(train_subj_labels)

train_feature_measurements <- read.csv ("UCI_HAR_Dataset/train/X_train.txt", sep = '', header = FALSE)
dim(train_feature_measurements)
# describe(train_feature_measurements)

# check that activity levels have same number of observations as measurements
print(ifelse((dim(train_feature_measurements)[1] != dim(train_act_labels)[1]) &
             (dim(train_feature_measurements)[1] != dim(train_subj_labels)[1]),
             "Error: length mismatch in train data",
             "Good: length match in train data"))
```


### Data transform

#### Update column names and add activity class and subject ID

``` {r}

# update the column names
colnames(test_act_labels) <- c("activity_class_encoding")
colnames(test_subj_labels) <- c("subject_id")
colnames(test_feature_measurements) = features$feature_name

colnames(train_act_labels) <- c("activity_class_encoding")
colnames(train_subj_labels) <- c("subject_id")
colnames(train_feature_measurements) = features$feature_name

# add activity names
test_feature_measurements$activity <- factor(test_act_labels$activity_class_encoding)
levels(test_feature_measurements$activity) <- act_labels$activity_name

train_feature_measurements$activity <- factor(train_act_labels$activity_class_encoding)
levels(train_feature_measurements$activity) <- act_labels$activity_name

# add subject ID (separate train id from test id)
test_feature_measurements$subjectid <- factor(test_subj_labels$subject_id)
train_feature_measurements$subjectid <- factor(train_subj_labels$subject_id)
``` 

#### extracts only the measurements on the mean and standard deviation

``` {r}

## use dplyr select to get a subset
#test_measurements_filtered <- select(test_feature_measurements, matches("-std"))
## Error: found duplicated column name:  

#columns <- grep("-(std|mean)", colnames(test_feature_measurements), value=TRUE)
#columns <- grep("-(std|mean)", colnames(test_feature_measurements), value=FALSE)
#select(test_feature_measurements, columns)

getDesiredColumnIndices <- function (df) {
  columns = grep("-(std|mean)", colnames(df), value=FALSE)
  columns = c(columns, grep("activity", colnames(df), value=FALSE))
  columns = c(columns, grep("subjectid", colnames(df), value=FALSE))
  columns
}

columns <- getDesiredColumnIndices(test_feature_measurements)
test_measurements_subset <- test_feature_measurements[,columns]

columns <- getDesiredColumnIndices(train_feature_measurements)
train_measurements_subset <- train_feature_measurements[,columns]

```


#### Merges the training and the test sets to create one data set.


``` {r}
# now merge the data frames
feature_measurements <- rbind(train_measurements_subset, 
                              test_measurements_subset)
dim(feature_measurements)
```



#### Create derived data set with the average of each variable for each activity and each subject.

``` {r}
tidy_feature_measurements <- feature_measurements %>%
  group_by(subjectid, activity) %>%
  summarise_each(funs(mean)) %>%
  ungroup() %>% 
  arrange()
```

##### Start of some EDA

``` {r Start of some EDA}
#describe(tidy_feature_measurements)

ggplot(data = subset(tidy_feature_measurements, activity == 'LAYING'),
       aes(x = tidy_feature_measurements$"tBodyGyro-mean()-X", y = tidy_feature_measurements$"tBodyGyro-mean()-Y") ) +
  geom_point()

ggplot(data = tidy_feature_measurements,
       aes(x = tidy_feature_measurements$"tGravityAcc-std()-Y")) +
  geom_histogram()

```


