---
title: "Udacity 651 PS5"
output:
  html_document:
    toc: true
    theme: united
---

Problem set 5
========================================================


``` {r Getting started}
library(ggplot2)
library(dplyr)
setwd("~/projects/Classes/FoundationsOfDataScience_sliderule/github/UD651/L5")

suppressMessages(library(dplyr))
data(diamonds)
```

***

### Price Histograms with Facet and Color

```{r Price Histograms with Facet and Color}
# Create a histogram of diamond prices.
# Facet the histogram by diamond color
# and use cut to color the histogram bars.

# The plot should look something like this.
# http://i.imgur.com/b5xyrOu.jpg

# Note: In the link, a color palette of type
# 'qual' was used to color the histogram using
# scale_fill_brewer(type = 'qual')

ggplot(aes(x = price),
       data = diamonds) +
  facet_wrap( ~ color) +
  geom_histogram(aes(color = cut)) 

```


Notes:

***

###  Price vs. Table Colored by Cut

```{r Price vs. Table Colored by Cut}
# Create a scatterplot of diamond price vs.
# table and color the points by the cut of
# the diamond.

# The plot should look something like this.
# http://i.imgur.com/rQF9jQr.jpg

# Note: In the link, a color palette of type
# 'qual' was used to color the scatterplot using
# scale_color_brewer(type = 'qual')

ggplot(aes(x = table, y = price),
       data = diamonds) +
  geom_point(aes(color = cut)) +
  scale_color_brewer(type = 'qual')

```



Notes:

***

###  Typical Table Value

Notes:

```{r  Typical Table Value}

```

Response:

typical table range for ideal cut:
53 - 57

typical table range for premium cut:
58 - 62



***

###  Price vs. Volume and Diamond Clarity

Notes:

```{r  Price vs. Volume and Diamond Clarity}
# Create a scatterplot of diamond price vs.
# volume (x * y * z) and color the points by
# the clarity of diamonds. Use scale on the y-axis
# to take the log10 of price. You should also
# omit the top 1% of diamond volumes from the plot.

# Note: Volume is a very rough approximation of
# a diamond's actual volume.

# The plot should look something like this.
# http://i.imgur.com/excUpea.jpg

# Note: In the link, a color palette of type
# 'div' was used to color the scatterplot using
# scale_color_brewer(type = 'div')
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
ninetynine_percentile <- quantile(diamonds$volume, 0.99)

ggplot(aes(x = volume, y = price),
       data = subset(diamonds, (volume < ninetynine_percentile) & (volume > 0))) +
  geom_point(aes(color = clarity)) +
  scale_y_log10() +
  scale_color_brewer(type = 'div') 
```

Response:



***

###  Proportion of Friendships Initiated


Notes:

```{r  Proportion of Friendships Initiated}
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')

# Many interesting variables are derived from two or more others.
# For example, we might wonder how much of a person's network on
# a service like Facebook the user actively initiated. Two users
# with the same degree (or number of friends) might be very
# different if one initiated most of those connections on the
# service, while the other initiated very few. So it could be
# useful to consider this proportion of existing friendships that
# the user initiated. This might be a good predictor of how active
# a user is compared with their peers, or other traits, such as
# personality (i.e., is this person an extrovert?).

# Your task is to create a new variable called 'prop_initiated'
# in the Pseudo-Facebook data set. The variable should contain
# the proportion of friendships that the user initiated.
pf$prop_initiated <- with(pf, friendships_initiated/friend_count)

#qplot(prop_initiated, data = pf)
```

Response:
***

###  prop_initiated vs. tenure

Notes:

```{r  prop_initiated vs. tenure}
# Create a line graph of the median proportion of
# friendships initiated ('prop_initiated') vs.
# tenure and color the line segment by
# year_joined.bucket.

# Recall, we created year_joined.bucket in Lesson 5
# by first creating year_joined from the variable tenure.
# Then, we used the cut function on year_joined to create
# four bins or cohorts of users.

# (2004, 2009]
# (2009, 2011]
# (2011, 2012]
# (2012, 2014]

# The plot should look something like this.
# http://i.imgur.com/vNjPtDh.jpg
# OR this
# http://i.imgur.com/IBN1ufQ.jpg

pf$year_joined <- with(pf,
                    2014 - ceiling(tenure/365))
pf$year_joined.bucket <- with(pf,
                           cut(year_joined,
  breaks = c(2004, 2009, 2011, 2012, 2014))
)

pf$prop_initiated.bucket <- with(pf, cut(year_joined,
  breaks = c(2004, 2009, 2011, 2012, 2014))
)

ggplot(aes(x = tenure, y = prop_initiated), 
       data = subset(pf, !is.na(prop_initiated) & !is.na(tenure) )) +
  geom_line(aes(color = year_joined.bucket), 
            stat = 'summary', fun.y = median)

```

Response:

***

###  Smoothing prop_initiated vs. tenure

Notes:

```{r  Smoothing prop_initiated vs. tenure}
ggplot(aes(x = tenure, y = prop_initiated), 
       data = subset(pf, !is.na(prop_initiated) & !is.na(tenure) )) +
  geom_smooth(aes(color = year_joined.bucket))

```

Response:

..which group initiated the greatest proportion of its facebook friendships?

group after 2012

for the group with highest initiated, what is mean proportion?
0.665

why do you think this groups proprtion is higher?

since they joined more recently, they would be likely to increase their number of friendships by requesting/initiating friends.  The longer a user has been on Facebook, the more likely it is that someone else has found them through searching and the other person initiated the friendship.

***

###  Price/Carat Binned, Faceted, & Colored

Notes:

```{r  Price/Carat Binned, Faceted, & Colored}
# Create a scatter plot of the price/carat ratio
# of diamonds. The variable x should be
# assigned to cut. The points should be colored
# by diamond color, and the plot should be
# faceted by clarity.

# The plot should look something like this.
# http://i.imgur.com/YzbWkHT.jpg.

# Note: In the link, a color palette of type
# 'div' was used to color the histogram using
# scale_color_brewer(type = 'div')

ggplot(aes(x = cut, y = price/carat),
       data = diamonds) +
  facet_wrap( ~ clarity) +
  geom_jitter(aes(color = color), alpha = 5/10, position = position_jitter(h = 0)) +
  scale_color_brewer(type = 'div')

```

Response:

