Lesson 5
========================================================


``` {r Getting started}
library(ggplot2)
library(dplyr)
setwd("~/projects/Classes/FoundationsOfDataScience_sliderule/github/UD651/L5")
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
suppressMessages(library(dplyr))
```

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
### does not plot; Error : Mapping a variable to y and also using stat="bin".
#ggplot(aes(x = gender, y = age),
#       data = subset(pf, !is.na(gender))) + geom_histogram()

ggplot(aes(x = gender, y = age), 
       data = subset(pf, !is.na(gender))) + geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4)

ggplot(aes(x = age, y = friend_count), 
       data = subset(pf, !is.na(gender))) + 
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)

```

***

### Third Qualitative Variable - exercise

``` {r Third Qualitative Variable - exercise}
# Write code to create a new data frame,
# called 'pf.fc_by_age_gender', that contains
# information on each age AND gender group.

# The data frame should contain the following variables:

#    mean_friend_count,
#    median_friend_count,
#    n (the number of users in each age and gender grouping)

# Here is an example of the structure of your data frame. Your
# data values will be different. Note that if you are grouping by
# more than one variable, you will probably need to call the
# ungroup() function. 

#   age gender mean_friend_count median_friend_count    n
# 1  13 female          247.2953                 150  207
# 2  13   male          184.2342                  61  265
# 3  14 female          329.1938                 245  834
# 4  14   male          157.1204                  88 1201

pf.fc_by_age_gender <- subset(pf, !is.na(gender)) %.%
  group_by(age, gender) %.%
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n = n()) %.%
  arrange(age) %.%
  ungroup()




names(pf.fc_by_age_gender)
head(pf.fc_by_age_gender)


```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}

ggplot(aes(x = age, y = median_friend_count), 
       data = pf.fc_by_age_gender) + 
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)
```

***

### Thinking in Ratios
Notes:

how many times more friends does F have than M?

***

### Wide and Long Format
Notes:

reshape data -> wide format from long format.

reshape2

***

### Reshaping Data
Notes:

```{r reshaping data}
#install.packages('reshape2')
library(reshape2)

pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')
head(pf.fc_by_age_gender.wide)

```


``` {r reshaping data - alternate}
#pf.fc_by_age_gender.wide <-     subset(pf.fc_by_age_gender[c('age', 'gender', #'median_friend_count')],                   !is.na(gender)) %>% 
#    spread(gender, median_friend_count) %>% 
#    mutate(ratio = male / female) 
```

***

### Ratio Plot
Notes:

```{r Ratio Plot}

# Plot the ratio of the female to male median
# friend counts using the data frame
# pf.fc_by_age_gender.wide.

# Think about what geom you should use.
# Add a horizontal line to the plot with
# a y intercept of 1, which will be the
# base line. Look up the documentation
# for geom_hline to do that. Use the parameter
# linetype in geom_hline to make the
# line dashed.

# The linetype parameter can take the values 0-6:
# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash

ggplot(aes(x = age, y = female / male), 
       data = pf.fc_by_age_gender.wide) + 
  geom_line() +
  geom_smooth() +
  geom_hline(aes (yintercept=1), alpha = 6/10, color = 'red', linetype = 2)
```

***

### Third Quantitative Variable
Notes:

tenure : days since joining FB
2014 is reference year
```{r Third Quantitative Variable}
# Create a variable called year_joined
# in the pf data frame using the variable
# tenure and 2014 as the reference year.

# The variable year joined should contain the year
# that a user joined facebook.

### question is ambiguous because it doesn't say whther to start windows at beginning or end of 2014

year_joined <- with(pf, 
                    2014 - trunc(tenure/365))

pf$year_joined <- with(pf, 
                    2014 - ceiling(tenure/365))  

```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
summary(pf$year_joined )
table(pf$year_joined )

# Create a new variable in the data frame
# called year_joined.bucket by using
# the cut function on the variable year_joined.

# You need to create the following buckets for the
# new variable, year_joined.bucket

#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]

# Note that a parenthesis means exclude the year and a
# bracket means include the year.

pf$year_joined.bucket <- with(pf, 
                           cut(year_joined, 
  breaks = c(2004, 2009, 2011, 2012, 2014))
)

```

***
`
### Plotting it All Together
Notes:

```{r Plotting it All Together}

ggplot(aes(x = year_joined.bucket),
       data = subset(pf, !is.na(gender))) + geom_histogram()

# Create a line graph of friend_count vs. age
# so that each year_joined.bucket is a line
# tracking the median user friend_count across
# age. This means you should have four different
# lines on your plot.

# You should subset the data to exclude the users
# whose year_joined.bucket is NA.

ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = median)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
# Write code to do the following:

# (1) Add another geom_line to code below
# to plot the grand mean of the friend count vs age.

# (2) Exclude any users whose year_joined.bucket is NA.

# (3) Use a different line type for the grand mean.

# As a reminder, the parameter linetype can take the values 0-6:

# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash
ggplot(aes(x = age, y = friend_count), 
              data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = mean, linetype = 2)

```

***

### Friending Rate
Notes:

```{r Friending Rate}

pf.fc_rate <- with(subset(pf, tenure >= 1),
                   friend_count / tenure)
summary(pf.fc_rate)


```

***

### Friendships Initiated
Notes:

What is the median friend rate?

0.2205
0.22

What is the maximum friend rate?

417.0000

```{r Friendships Initiated}
# Create a line graph of mean of friendships_initiated per day (of tenure)
# vs. tenure colored by year_joined.bucket.

# You need to make use of the variables tenure,
# friendships_initiated, and year_joined.bucket.

# You also need to subset the data to only consider user with at least
# one day of tenure.

ggplot(aes(x = -tenure, y = friendships_initiated / tenure), 
              data = subset(pf, tenure >= 1)) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean)

```


***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

```{r Bias-Variance Tradeoff Revisited smooth}
# Instead of geom_line(), use geom_smooth() to add a smoother to the plot.
# You can use the defaults for geom_smooth() but do color the line
# by year_joined.bucket
ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Sean's NFL Fan Sentiment Study
Notes:

spline <- bad model
moving average <- 1 day, 7 day MA much better
***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo = read.csv('yogurt.csv')
# str(yo)

# change the id (household ID) from an int to a factor
yo$id <- factor(yo$id)
str(yo)

ggplot(aes(x = price), data = yo) + geom_histogram()
ggplot(aes(x = price), data = yo) + geom_histogram( binwidth=10)

```

What do you notice?

- distribution trends noticeable at different price points.

***

### Number of Purchases
Notes:

```{r Number of Purchases}
# Create a new variable called all.purchases,
# which gives the total counts of yogurt for
# each observation or household.

# One way to do this is using the transform
# function. You can look up the function transform
# and run the examples of code at the bottom of the
# documentation to figure out what it does.

# The transform function produces a data frame
# so if you use it then save the result to 'yo'!

# OR you can figure out another way to create the
# variable.

yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
#head(yo[3:10], 30)
```

***

### Prices over Time
Notes:

```{r Prices over Time}

#ggplot(aes(x = time, y = price), data = yo) + 
#  geom_point(alpha = 1/10, color = 'red')

ggplot(aes(x = time, y = price), data = yo) + 
  geom_jitter(alpha = 2/10, shape = 21, color = 'red')
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
# set the seed for reproducible resuults
set.seed(4563)
sample.ids <- sample(levels(yo$id), 16)

ggplot(aes(x = time, y = price), 
       data = subset(yo, id %in% sample.ids)) + 
  facet_wrap( ~ id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)

ggsave('yogurt_sample16.png')

```

I observed a variety of different purchasing patterns. Some price curves look like W's, with some heavy-consuming households.  There were many lighter households in the sample as well.  Most, but not all, households exhibited trends of increasing prices over time.

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix

``` {r Scatterplot Matrices}
#install.packages('GGally')
library(GGally)
theme_set(theme_minimal(20))

# seed 
set.seed(1836)
pf_subset <- pf [, c(2:15)]
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])

```
Notes:

Response:

CC of friendship initiated and friend count?
0.769

CC of age and mobile likes?
-0.0475

What type of plot below the diagonal?  What type of plots for gender pairs?
scatter plot.  histogram


***

### Even More Variables
Notes:

***

wget https://s3.amazonaws.com/udacity-hosted-downloads/ud651/nci.tsv

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

Learned some on how to transform variables and generate relative variable for additional graphing techniques.  Learned how to bucket variables for analysis and plotting, as well as categorizing factors and plotting conditional summaries.

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

