---
title: "Data Wrangling with R tutorial"
output:
  html_document:
    toc: true
    theme: united
---

## Data Wrangling with R


https://ramnathv.github.io/pycon2014-r/explore/README.html


*** 

### Getting started

```{r libraries and wd}
#pkgs <- c('reshape2', 'plyr', 'ggplot2', 'dplyr', 'data.table', 'Lahman')
#install.packages(pkgs)
#install.packages('data.table')
#install.packages('Lahman')
#install.packages("tidyr")

library(reshape2)
library(plyr)
library(ggplot2)
library(dplyr)
library(data.table)
library(Lahman)

library(tidyr)

setwd("~/projects/Classes/FoundationsOfDataScience_sliderule/github/part4_data_wrangling/tutorial")

```

*** 

### Explore messy Data

https://ramnathv.github.io/pycon2014-r/explore/tidy.html

```{r pew pew pew}

pew <- read.delim(
#  file = "http://stat405.had.co.nz/data/pew.txt",
  file = "pew.txt",
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = F
)
head(pew)


tb <- read.csv(
#  file = "http://stat405.had.co.nz/data/tb.csv",
  file = "tb.csv",
  header = TRUE, 
  stringsAsFactors = FALSE
)
head(tb)

weather <- read.delim(
# file = "http://stat405.had.co.nz/data/weather.txt",
 file = "weather.txt",
 stringsAsFactors = FALSE
)
head(weather)

```

#### tb

Except for `iso2` and `year`, the rest of the columns headers are actually values of a lurking variable, in fact combination of **two** lurking variables, `gender` and `age`.

#### weather
This dataset seems to have two problems. First, it has variables in the rows in the column `element`. Second, it has a variable `d` in the column header spread across multiple columns.

``` {r}
# # head(df)
# colnames(df) <- c("country", 1980:2007)
# #str(df)
# ldf <- gather(df, "year", "percentage", 2:29)
# 
# # want the years to be numeric instead of factor?
# ldf$year_numeric <- with(ldf, as.numeric(levels(year))[year])
# # head(ldf) 
# str(ldf)
```

#### Causes of Messiness

There are various features of messy data that one can observe in practice. Here are some of the more commonly observed patterns.

- Column headers are values, not variable names
- Multiple variables are stored in one column
- Variables are stored in both rows and columns
- Multiple types of experimental unit stored in the same table
- One type of experimental unit stored in multiple tables



*** 

### Reshape


```{r melt the pew income data}
#library(reshape2)
pew_tidy <- melt(
  data = pew,
  id = "religion",
  variable.name = "income",
  value.name = "frequency"
)
# head(pew_tidy)
```

```{r melt similarly the tb data}
#library(reshape2)

# set column `new_sp` to NULL and clean up column names
tb$new_sp = NULL
names(tb) <- gsub("new_sp_", "", names(tb))

# Use na.rm = TRUE to remove missing observations
tb_tidy <- melt(
  data = tb,
  id = c("iso2", "year"),
  variable.name = "gender_age",
  value.name = "cases",
  na.rm = TRUE
)
#head(tb_tidy)


# use regex substitutions to split gender_age into gender and age group
library(plyr)
tb_tidy <- mutate(tb_tidy,
  gender = sub("^([m|f])(.*)$", "\\1", gender_age),
  age = sub("^([m|f])(.*)$", "\\2", gender_age),
  gender_age = NULL
)
tb_tidy <- tb_tidy[c('iso2', 'year', 'gender', 'age', 'cases')]
#head(tb_tidy)

```



```{r weather data needs some tweaking too}
library(reshape2)

# first melt the data and convert the day variable into numbers.
weather_tidy <- melt(
  data = weather,  
  id = 1:4,
  variable.name = "day",   # d1-d31
  value.name = "temperature",
  na.rm = TRUE
)

# trim the /^d/ from the day column
weather_tidy <- mutate(weather_tidy,
  day = sub("^d", "", day)  
)

weather_tidy2 <- dcast(
  data = weather_tidy,
  formula = id + year + month + day ~ element, 
  value.var = "temperature"
)

head(weather_tidy)
head(weather_tidy2)


```

#### billboards data

``` {r tidy up billboards}
billboards <- read.csv(
#  file = "http://stat405.had.co.nz/data/billboard.csv",
  file = "billboard.csv",
  stringsAsFactors = FALSE
)
names(billboards) <- gsub("\\.", "_", names(billboards))

# some more cleanup

# char encodings
billboards <- mutate(billboards,
  artist_inverted = iconv(artist_inverted, from = "MAC", to = "UTF-8")
)

# melt the x67th_week ... columns
billboards_tidy <- melt(billboards, 
  id = 1:7,
  variable.name = "week",
  value.name = "rank",
  na.rm = TRUE
)

# now tweak the week format from 'x1st_week' to '1'
billboards_tidy <- mutate(billboards_tidy,
  week = as.numeric(gsub("^x([[:digit:]]+).*", "\\1", week))  
)


```

*** 

### Summarize


```{r baby names dataset}
# https://ramnathv.github.io/pycon2014-r/explore/data/bnames2.csv.bz2
bnames2 <- read.csv("bnames2.csv.bz2")
```

#### summarize specific names

```{r gregs}
greg <- subset(bnames2, name == "Greg")

#library(ggplot2)
qplot(x = year, y = prop, data = greg, geom = 'line')

```


```{r michelles}
michelle <- subset(bnames2, name == "Michelle")

#library(ggplot2)
qplot(x = year, y = prop, data = michelle, geom = 'line')
qplot(x = year, y = prop, data = michelle, geom = 'point')

### sawtooth in output, because of two genders ('sex')
qplot(x = year, y = prop, data = michelle, geom = 'line', group = sex)
```



```{r davids}
david <- subset(bnames2, name == "David")
qplot(x = year, y = prop, data = david, geom = 'line', group = sex)

```

```{r svetas and ilonas}
ilona <- subset(bnames2, name == "Ilona")
qplot(x = year, y = prop, data = ilona, geom = 'point')
```

#### soundex

``` {r greg soundex}
greg_soundex = greg$soundex[1]
greg_like <- subset(bnames2, soundex == greg_soundex)
qplot(x = year, y = prop, data = greg_like, geom = 'point')
```

``` {r ilona soundex}
ilona_soundex = ilona$soundex[1]
ilona_like <- subset(bnames2, soundex == ilona_soundex)
qplot(x = year, y = prop, data = ilona_like, geom = 'point')
#qplot(x = year, y = prop, data = ilona_like, geom = 'line')
```

### data manipulation verbs

https://ramnathv.github.io/pycon2014-r/explore/data/births.csv

``` {r births data set}
births <- read.csv("births.csv")
qplot(year, births, data = births, color = sex, geom = 'line')

```

#### Join

``` {r join birth with baby names}
bnames2_b <- join(bnames2, births, by = c("sex", "year"))
```

``` {r now get full counts for greg}
greg <- subset(bnames2_b, name == 'Greg')
greg <- mutate(greg, tot = prop * births)
qplot(year, births, data = greg, geom = 'line')
```

### what is the most popular name in the US across all the years (1880 - 2008)?

``` {r greg summarized}
greg_tot <- summarize(greg, tot = sum(prop * births))
```


How do we do this for all names? In terms of pseudocode, here is what we would do.

- Split the dataset by name.
- Apply the summary computations to each name.
= Combine the summaries created into a single dataset.


``` {r split - apply - combine}
# Split
pieces <- split(bnames2_b, bnames2_b$name)

# Apply
results <- vector("list", length(pieces))
for (i in seq_along(pieces)){
  results[[i]] <- summarize(pieces[[i]],
     name = name[1],
     tot = sum(prop * births)
  )
}

# Combine
result <- do.call("rbind", results)
```

``` {r now sort them}
most_pop_name <- arrange(result, desc(tot))[1,"name"]
most_pop_name
```

***


## Split-Apply-Combine: I

```{r aggregate is another way}
bnames2_b = mutate(bnames2_b, tot = prop * births)
result <- aggregate(formula = tot ~ name, data = bnames2_b, FUN = sum)
```


### most popular name redux

```{r popular name another way}
result2 <- aggregate(formula = tot ~ name + sex, data = bnames2_b, FUN = sum, 
    subset = (year >= 2000))
most_pop_boy <- arrange(subset(result2, sex == "boy"), desc(tot))[1, "name"]
most_pop_girl <- arrange(subset(result2, sex == "girl"), desc(tot))[1, "name"]
```

### some operators

#### apply

``` {r}
m <- matrix(c(1:10, 11:20), nrow = 10, ncol = 2)
# 1 is the row index 2 is the column index
apply(m, 1, sum)

apply(m, 2, sum)

apply(m, 1, mean)
apply(m, 2, mean)
```

#### lapply - list apply, sapply (list of matrix) apply, mapply - map apply, tapply - subsets of a vector

``` {r}
my_list <- list(a = 1:10, b = 2:20)
lapply(my_list, mean)

my_list
x <- sapply(my_list, mean)
x

class(x)

list_1 <- list(a = c(1:10), b = c(11:20))
list_1

list_2 <- list(c = c(21:30), d = c(31:40))
list_2

mapply(sum, list_1$a, list_1$b, list_2$c, list_2$d)

head(warpbreaks)


with(warpbreaks, tapply(breaks, list(wool, tension), mean))

```

#### by - functions to subset of data frame

``` {r by and by}

head(iris)

by(iris[, 1:2], iris[, "Species"], summary)

by(iris[, 1:2], iris[, "Species"], sum)
```


#### replicate - generate datasets with distributions

``` {r}
replicate(10, rnorm(10))
replicate(10, rnorm(10), simplify = TRUE)
```


### Split-Apply-Combine: II




``` {r plyr}
library(plyr)
result <- ddply(
  .data = subset(bnames2_b, year >= 2000),
  .variables = c('sex', 'name'),
  .fun = function(p){
    summarize(p, tot = sum(tot))
  }
)

resultB <- ddply(subset(bnames2_b, year >= 2000), 
  .(sex, name), summarize, tot = sum(tot)
)
```

``` {r baseball rbis}
head(baseball[,1:16])


rbi_vs_year <- function(df){
  df <- mutate(df, year = year - min(year))
  lm(rbi ~ year, data = df)
}

models <- dlply(baseball, .(id), rbi_vs_year)
models[[1]]

coefs <- ldply(models, coef)
qplot(`(Intercept)`, year, data = coefs)
```


``` {r more baseball}
games <- ddply(Batting, .(playerID), summarize, total = sum(G))
head(arrange(games, desc(total)), 5)

library(dplyr)
games_d <- Batting %.%
  group_by(playerID) %.%
  summarize(total = sum(G)) %.%
  arrange(desc(total)) %.%
  head(5)


library(data.table)
Batting_DT <- data.table(Batting)
games_dt <- Batting_DT[,
  list(total = sum(G)),
  "playerID"
][
  head(order(-total), 5),
]
```

