---
title: "Udacity 651 PS4"
output:
  html_document:
    toc: true
    theme: united
---

Problem set 4

*** 

Getting started

```{r libraries and wd}
library(ggplot2)
setwd("~/projects/Classes/FoundationsOfDataScience_sliderule/github/UD651/L4")
data(diamonds)
```

*** 

```{r Scatterplot of diamond}
ggplot(aes(x = x, y = price), data = diamonds) + 
  geom_jitter(alpha = 1/10)
```


#### What are your observations about the scatterplot of price vs. x?
Response:

there is a price gap around x = 5, and price is positively correlated with x. 
there seems to be an exponential relationship of price vs x


*** 


``` {r relations of price vs x,y,z}
library(gridExtra)
p1 = ggplot(aes(x = x, y = price), data = diamonds) + 
  geom_jitter(alpha = 1/10)

p2 = ggplot(aes(x = y, y = price), data = diamonds) + 
  geom_jitter(alpha = 1/20)

p3 = ggplot(aes(x = z, y = price), data = diamonds) + 
  geom_jitter(alpha = 1/20)
  
grid.arrange(p1, p2, p3, ncol = 1)
```


``` {r Correlations of price vs x,y,z}

with(diamonds, cor.test(x, price))
with(diamonds, cor.test(y, price))
with(diamonds, cor.test(z, price))

```

#### What is the correlation ...


#### between price and x?
Response: 

positive, exponential
0.88

####  between price and y?
Response:

positive, exponential
0.87

####  between price and z?
Response:

positive, exponential
0.86


***

### price vs. depth

``` {r Scatterplot of price vs depth}
ggplot(aes(x = depth, y = price), data = diamonds) + 
  geom_jitter(alpha = 3/10)
```

#### Overlapping of points


``` {r Transparency and scale}
# Change the code to make the transparency of the
# points to be 1/100 of what they are now and mark
# the x-axis every 2 units. 

ggplot(data = diamonds, aes(x = depth, y = price)) + 
  geom_point(alpha = 1/100) +
  scale_x_continuous(breaks = seq(0, 80, 2))
```



