Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
library(ggplot2)
setwd("~/projects/Classes/FoundationsOfDataScience_sliderule/github/UD651/L4")
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
#qplot(x= age, y= friend_count, data = pf)
qplot(age, friend_count, data = pf)
```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20, position = position_jitter(h = 0)) +
  coord_trans(y="sqrt") +
  xlim(13,90)

```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) +
  xlim(13,90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20) +
  xlim(13,90)
  
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
qplot(age, friend_count, data = pf) +
 coord_trans(ytrans="sqrt")

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) +
  coord_trans(y="sqrt") +
  xlim(13,90)

summary(pf$age)

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friendships_initiated + 1), data = pf) + 
  geom_jitter(alpha = 1/20, position = position_jitter(h = 0)) +
  coord_trans(y="log10") +
  xlim(13,90)

```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
#install.packages('dplyr')
library(dplyr)

age_groups <- group_by(pf, age)
pf.fc_by_age <- summarize(age_groups, 
        friend_count_mean = mean(friend_count),
        friend_count_median = median(friend_count),
        n = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age)

head(pf.fc_by_age)


```

```{r Conditional Means Alternate Code}
library(dplyr)

pf.fc_by_age <- pf %.%
  group_by(age) %.%
  summarise(friend_count_mean = mean(friend_count),
        friend_count_median = median(friend_count),
        n = n()) %.%
   arrange(age)


head(pf.fc_by_age, 20)
```

Create your plot!

```{r Conditional Means Plot}
#ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) +
#  geom_line()
ggplot(aes(x = age, y = friend_count_median), data = pf.fc_by_age) +
  geom_line()
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count + 1), data = pf) + 
  coord_cartesian(xlim = c(13, 90)) +
  geom_jitter(alpha = 1/20, position = position_jitter(h = 0)) +
  coord_trans(y="log10")

ggplot(aes(x = age, y = friend_count), data = pf) + 
  coord_cartesian(xlim = c(13, 70), ylim = c(0,1000)) +
  geom_point(alpha = 1/20, 
             position = position_jitter(h = 0),
             color = 'orange') +
  geom_line(stat = 'summary', fun.y = median) +
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .1,
            linetype = 2, color = 'purple') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .9,
            linetype = 2, color = 'purple') 

```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
?cor.test

cor.test(pf$age, pf$friend_count,
         method=c("pearson"))
with(pf, cor.test(age, friend_count, method="pearson"))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

-0.027

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
#with(                 , cor.test(age, friend_count))
with(subset(pf, age <= 70), cor.test(age, friend_count,
                                   method="pearson"))
```

***

### Correlation Methods
Notes:

```{r Spearman correlation}
with(subset(pf, age <= 70), cor.test(age, friend_count,
                                   method="spearman"))
```

***

## Create Scatterplots
Notes:

```{r}
names(pf)
#qplot(www_likes_received, www_likes_received/likes_received, data = pf)
#qplot(www_likes_received, likes_received, data = pf) +
#  coord_cartesian(xlim = c(0, 12500), ylim = c(0,50000)) 
#qplot(www_likes_received, likes_received, data = pf) +
#  coord_cartesian(xlim = c(0, 2500), ylim = c(0,20000)) 

ggplot(aes(x = www_likes_received, y = likes_received), data = pf) + 
  coord_cartesian(xlim = c(0, 300), ylim = c(0,1250)) +
  geom_point(alpha = 1/10, 
             position = position_jitter(h = 0),
             color = 'blue') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(y = pf$www_likes_received)

```
***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) + 
  coord_cartesian(xlim = c(0, quantile(pf$www_likes_received, 0.95)),
                  ylim = c(0, quantile(pf$likes_received, 0.95))) +
  geom_point(alpha = 1/10, 
             position = position_jitter(h = 0),
             color = 'blue') +
  geom_smooth(method = "lm", color = "red") +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(y = pf$www_likes_received)
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calculation}
with(pf, cor.test(www_likes_received, likes_received, 
                  method="pearson"))
```

Response:
      cor 
0.9479902 

0.948

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
#install.packages('alr3')
library(alr3)
```

Create your plot!

```{r Temp vs Month}

```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}

```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}

```

***

### Age with Months Means

```{r Age with Months Means}

```

Programming Assignment
```{r Programming Assignment}

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

