---
title: "Udacity 651 PS4"
output:
  html_document:
    toc: true
    theme: united
---

Problem set 4

*** 

Getting started

```{r libraries and wd}
library(ggplot2)
library(dplyr)
setwd("~/projects/Classes/FoundationsOfDataScience_sliderule/github/UD651/L4")
data(diamonds)
```

*** 

```{r Scatterplot of diamond}
ggplot(aes(x = x, y = price), data = diamonds) + 
  geom_jitter(alpha = 1/10)
```


#### What are your observations about the scatterplot of price vs. x?
Response:

there is a price gap around x = 5, and price is positively correlated with x. 
there seems to be an exponential relationship of price vs x


*** 


``` {r relations of price vs x,y,z}
library(gridExtra)
p1 = ggplot(aes(x = x, y = price), data = diamonds) + 
  geom_jitter(alpha = 1/10)

p2 = ggplot(aes(x = y, y = price), data = diamonds) + 
  geom_jitter(alpha = 1/20)

p3 = ggplot(aes(x = z, y = price), data = diamonds) + 
  geom_jitter(alpha = 1/20)
  
grid.arrange(p1, p2, p3, ncol = 1)
```


``` {r Correlations of price vs x,y,z}

with(diamonds, cor.test(x, price))
with(diamonds, cor.test(y, price))
with(diamonds, cor.test(z, price))

```

#### What is the correlation ...


#### between price and x?
Response: 

positive, exponential
0.88

####  between price and y?
Response:

positive, exponential
0.87

####  between price and z?
Response:

positive, exponential
0.86


***

### price vs. depth

``` {r Scatterplot of price vs depth}
ggplot(aes(x = depth, y = price), data = diamonds) + 
  geom_jitter(alpha = 3/10)
```

### Overlapping of points


``` {r Transparency and scale}
# Change the code to make the transparency of the
# points to be 1/100 of what they are now and mark
# the x-axis every 2 units. 

ggplot(data = diamonds, aes(x = depth, y = price)) + 
  geom_point(alpha = 1/100) +
  scale_x_continuous(breaks = seq(0, 80, 2))
```

#### Based on the scatterplot of depth vs. price, most diamonds are betweeen what values of depth?

Response:
lower limit to upper limit

59  to 65
59  to 64  ACCEPTED

``` {r correlation of depth vs price}
with(diamonds, cor.test(depth, price))
```

#### What is the correlation of depth vs. price

Response:

-0.010

#### Based on the correlation coefficient, would you use depth to predict the price of a diamond?  and why?

Response:

No


much smaller than |0.3|

***

### price vs. carat

``` {r price vs carat}
# Create a scatterplot of price vs carat
# and omit the top 1% of price and carat
# values.

ggplot(data = diamonds, aes(x = carat, y = price)) + 
  geom_point(alpha = 1/100) +
  coord_cartesian(xlim = c(0, quantile(diamonds$carat, 0.99)),
                  ylim = c(0, quantile(diamonds$price, 0.99)))
```



***

### price vs. volume

``` {r price vs volume}
# Create a scatterplot of price vs. volume (x * y * z).
# This is a very rough approximation for a diamond's volume.

# Create a new variable for volume in the diamonds data frame.
# This will be useful in a later exercise.

ggplot(data = diamonds, aes(x = x * y * z, y = price)) + 
  geom_point(alpha = 1/10) +
  coord_trans(y="log10") +
  coord_cartesian(xlim = c(0, 354))
# 354 is 99% bound of volume

diamonds$volume <- diamonds$x * diamonds$y * diamonds$z

```


#### What are your observations from the price vs. volume scatterplot?
Response:

some outliers. looks exponential. high upper range  in bounding box at higher volumes



``` {r correlation of depth vs volume}
with(subset(diamonds, (volume != 0) & (volume < 800)) , 
            cor.test(price, volume))
```

#### What is the correlation of price and volume? 

Exclude diamonds that ahve a volume of 0 or that are greater than or equal to 800.


Response:

0.92

***

### Adjustments - price vs. volume

``` {r Adjustments - price vs. volume}
# Subset the data to exclude diamonds with a volume
# greater than or equal to 800. Also, exclude diamonds
# with a volume of 0. Adjust the transparency of the
# points and add a linear model to the plot. (See the
# Instructor Notes or look up the documentation of
# geom_smooth() for more details about smoothers.)

# We encourage you to think about this next question and
# to post your thoughts in the discussion section.

# Do you think this would be a useful model to estimate
# the price of diamonds? Why or why not?

ggplot(data = subset(diamonds, (volume != 0) & (volume < 800)), 
       aes(x = x * y * z, y = price)) + 
  geom_point(alpha = 1/20) +
  stat_smooth(method = "lm", color = "red") +
  coord_cartesian(xlim = c(1, 354))

# probably move away from a linear model, since data looks to be more exponential

```


***

### Mean Price by Clarity

``` {r Mean Price by Clarity}
# Use the function dplyr package
# to create a new data frame containing
# info on diamonds by clarity.

# Name the data frame diamondsByClarity

# The data frame should contain the following
# variables in this order.

#       (1) mean_price
#       (2) median_price
#       (3) min_price
#       (4) max_price
#       (5) n

# where n is the number of diamonds in each
# level of clarity.
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
data(diamonds)

diamondsByClarity <- diamonds %.%
  group_by(clarity) %.%
  summarise(mean_price = mean(price),
            median_price = median(price),
            min_price = min(price),
            max_price = max(price),
            n = n()) %.%
   arrange(clarity)

```


***

###  Bar Charts of Mean Price

``` {r Bar Charts of Mean Price}


```
